The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-06 VegOzone.tex
Finished large parts of section 3.1. Missing ozone climatology and spell check.

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}
\begin{itemize}
\item ozone in general
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

In our assessment on ozone damage on vegetation in the subarctic in relation to the 2018 heatwave, we focus on the NIBIO operated local site Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley (Section~\ref{subsec:svanhovd}). Throughout the paper, we will use Svanvik synonymously for the atmospheric monitoring site at Svanhovd. The available meteorological monitoring data are described in Section~\ref{subsubsec:atmo_svanvik}. In Section~\ref{subsubsec:ozone_garden}, the ozone garden -- a plant-based ozone monitoring -- will be described in more detail. Other data used for evaluation of the ozone data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use as well reanalysis products (Section~\ref{subsubsec:ozone_rea}) as complement when observational data was not available.

@@

The locations of the atmospheric monitoring site as well as the ozone garden with respect to the main buildings of the NIBIO research station are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}. 
\emph{Comment: Invite co-authors from NILU and NIBIO?}
{\bf TODO: Some more general information about the research station? E.g., what NIBIO us currently researching there, that it also hosts a conference site, and national park center.}

@@

Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured in 2019, for data acquisition started later than in 2018. In 2018, high summer ozone concentrations ($\chem{[O_3]}> 40\,\unit{ppb}$) were observed in July. A part of these peak values may, however, coincident with the time period of missing data. In the summer of 2019, ozone concentrations stayed below the threshold of $40\,\unit{ppb}$.

@@

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we have to derive a surface ozone concentration climatology at Svanvik. From 1986--1996 ozone monitoring has been operated by NILU at Svanhovd, but no ozone measurements have been conducted for the past $14\,\unit{years}$. Due to probabel changes in tropospheric background ozone and a reduction of summertime peak values related to precursor that fell under airquality regulations, we also have to have a look at ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted atmospheric monitoring in past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data is descibed in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which is not available at all times and spatially sparse, we also look at verious ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observation are locate on higher ground than Svanhovd which will result in higher surface ozone concentrations compared to data taken at Svanvik as ozone abundance decreases with altitude in the absence of local precursors. {\bf TODO: The following has to be confirmed and ellaborated on!} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

Data from before the 1990s is available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care {\bf TODO: citation NILU ozone report 200x}. Especially, ozone monitors did not undergo a regular re-calibration. This leads likely to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul had been relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have intruduced a shift in the data series denotes as Jergul/Karasjok in the following. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlapp of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas herin indicate the time periods with insufficient quality control as mentioned above.

@@

Details of renalysis data sets refered to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone (MACC and CAMSRA). The temporal as well as spatial resolution of these reanalysis products is rather corse: 3-hourly and $0.75^\circ\times 0.75^\circ$ or roughly $(78.1\times 83.4)\,\unit{km}$ at the location of Svanvik, respectively. In addition, higher resolution surface ozone reanalysis for whole Europe is available from Copernicus service for regional air quality control. The regional model reanalysis ensemble mean is at fairly high spatial and temporal resolution compared to the data from ECMWF: $0.1^\circ\times 0.1^\circ$ or roughly $(10.4\times 11.1)\,\unit{km}$ and 1-hourly, respectively. The covered time periods differ and no data are available for before the turn of the millenium. The CAMSRA and CAMS Regional Air Quality ensemble mean are available in near realtime, but only former covers a period of sufficient length for climate analysis (2003 -- present). Their quality with respect to observations will be discussed in Section~\ref{subsubsec:clim_ozone}. Without too much foreshadowing, the results prevented us from fetching the whole CAMSRA data.

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
\begin{figure}[t]
  \includegraphics[width=8.3cm]{climatediagram}
  \caption{Climate diagram following Walter and Lieth. The diagram is based on climatological data for Svanvik/Pasvik (1992--2017) downloaded from LandbruksMeteorologiske Tjeneste. Monthly average temperatures (red line) are displayed with standard error of mean as error band. Average monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star.}
  \label{fig:climatediagram}
\end{figure}

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}

\begin{figure}[t]
  \includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs}
  \caption{Multi-annual mean of daily mean ozone at observation sites in Fennoscandia. The large spread in case of Svanvik and Janiskoski is due to the lower statistics. All northern Fennoscandic climatologies peak in spring, with peak values in April. In southern Norway, peak values appear only in May. The annual average ozone concentration \chem{\left<[O_3]\right>} at Svanvik is $6.6\,\unit{ppb}$ lower than at the sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN).}
  \label{fig:ozone_climatology_fenoscandic_obs}
\end{figure}

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{ozone_climatologies}
  \caption{Comparison of climatologies (northern Fennoscandia -- black and southern Fennoscandia -- grey) derived from observational data (EBAS, see Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}) with various ozone reanalysis products: (a) ECMWF: MACC reanalysis; (b) ECMWF: CAMS reanalysis; (c) Copernicus: Regional model reanalysis ensemble mean. As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches quite well in southern Fennoscandia but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in both northern and southern Fennoscandia, though the ozone concentrations in the north are slightly to low.}
  \label{fig:ozone_climatologies}
\end{figure*}

@@

\appendixfigures  %% needs to be added in front of appendix figures
\begin{figure}[t]
  \includegraphics[width=8.3cm]{greening_season_change_Svanvik}
  \caption{Estimated shift and prolongation of greening season at Svanhovd over the past two decades based on the method and data described in \citet{GMD:Falk2019}.}
  \label{fig:greening_season_change_Svanvik}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_clim}
  \caption{Observed global irradiance at Svanvik. The grey lines indicate the $1\,\sigma$ level deducted from the climatology. (a) Climatology (hourly maxima, mean, and minima); deviation from climatology (b) 2018; (c) 2019.}
  \label{fig:global_rad_clim}
\end{figure}

>>>>

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}
\begin{itemize}
\item ozone in general
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item metrices of ozone: aot40 and pod
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

In our assessment of ozone damage on vegetation in the subarctic in relation to the 2018 heatwave, we focus on the NIBIO operated local site Svanhovd which lies close by the settlement of Svanvik in the Pasvik valley (Section~\ref{subsec:svanhovd}). Throughout the paper, we will use Svanvik synonymously for the atmospheric monitoring site at Svanhovd. The available meteorological monitoring data are described in Section~\ref{subsubsec:atmo_svanvik}. In Section~\ref{subsubsec:ozone_garden}, the ozone garden -- a plant-based ozone monitoring -- will be described in more detail. Other data used for evaluation of the ozone data taken at Svanhovd is described in Section~\ref{subsec:other_data}. We compare with observational data from other monitoring sites in northern Fennoscandia (Section~\ref{subsubsec:ebas}) and use as well reanalysis products (Section~\ref{subsubsec:ozone_rea}) as complement when observational data was not available.

@@

The locations of the atmospheric monitoring site as well as the ozone garden with respect to the main buildings of the NIBIO research station are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}. 
\emph{Comment: Invite co-authors from NILU and NIBIO?}
     {\bf TODO: Some more general information about the research station? E.g., what NIBIO us currently researching there, that it also hosts a conference center and national park center.}
     "The close proximity to Russia was a dominant factor when the new centre was built in 1992 and established as Svanhovd Environment Centre. The centre building has later been extended and modernised. It contains offices, a conference centre, a national park centre, a brown bear exhibition and a modern laboratory which is also accredited for DNA analysis of brown bear. The centre is surrounded by a botanical garden, one of three such gardens open to the public north of the Polar Circle.

     NIBIO Svanhovd in Pasvik valley is strongly committed to environmental research cooperation in the Barents region. The station disseminates knowledge and performs R\&D based upon the natural environment of this region. It also runs an up-to date DNA-laboratory, a national park centre, a botanical garden and a conference centre. Svanhovd has extensive cooperation with all countries in the Barents region."  - \href{https://www.nibio.no/en/about-eng/addresses/northern-norway/svanhovd}{NIBIO webpage, accessed April 2020}.\\

@@

Ozone concentrations \chem{[O_3]} measured in $2\,\unit{m}$ height above ground are averaged hourly. As can be seen in Fig.~\ref{fig:data_svanvik_2018_2019}a), \chem{[O_3]} peaks in spring (April/May) and reaches its minimum in late summer (July/August). The spring peak has not been captured completely in 2019, for data acquisition started later than in 2018. In 2018, high summer ozone concentrations ($\chem{[O_3]}> 40\,\unit{ppb}$) were observed in July. A part of these peak values may, however, coincident with the time period of missing data. In the summer of 2019, ozone concentrations stayed below the threshold of $40\,\unit{ppb}$ most of the times.

@@

\subsection{Other data}
\label{subsec:other_data}
To determine the significance of the ozone conditions in 2018, we have to derive a surface ozone concentration climatology for Svanvik. From 1986--1996 ozone monitoring has been operated by NILU at Svanhovd, but no ozone measurements have been conducted for the past $14\,\unit{years}$. Due to probabel changes in northern hemispheric tropospheric background ozone and a reduction of summertime peak values related to ozone precursors that fell under airquality regulations {\bf TODO: Citations!}, we also have to have a look at ozone observations at other sites in northern Fennoscandia. In regard to ozone, available observational data are sparse in this region. An overview over the geographical locations of conducted atmospheric monitoring in the past and present is given in Fig.~\ref{fig:station_map_fennoscandia}. The data is descibed in Section~\ref{subsubsec:ebas}. 
As additional complement to observational data which is not available at all times and spatially sparse, we also look at verious ozone reanalysis products (Section~\ref{subsubsec:ozone_rea}).

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observation are locate on higher ground than Svanhovd which will result in higher surface ozone concentrations compared to data taken at Svanvik as ozone abundance decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation!} {\bf TODO: The following has to be confirmed and ellaborated on!} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

Data from before the 1990s is available solely from Svanvik and Karasjok. These data, however, do not follow the high quality standard procedures implemented nowadays and have to be treated with care {\bf TODO: citation NILU ozone report 200x}. Especially, ozone monitors did not undergo a regular re-calibration. This leads likely to drifts in the observed data and may impose a false periodicity. In 1997, the ozone monitoring at Jergul was relocated downstream the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have intruduced a shift in the data series denotes as Jergul/Karasjok in the following. Only few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration. As can be seen in Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}, there is an overlapp of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik. The hatched areas herin indicate the time periods with insufficient quality control as mentioned above.

@@

Details of renalysis data sets refered to in this work are listed in Table~\ref{tab:ozone_rea}.
There are two global reanalysis products available at the European Centre for Medium-Range Weather Forecasts (ECMWF) that include atmospheric tracers, in particular ozone (MACC and CAMSRA). The temporal as well as spatial resolution of these reanalysis products is rather coarse: 3-hourly and $0.75^\circ\times 0.75^\circ$ or roughly $78.1\,\unit{km}\times 83.4\,\unit{km}$ at the location of Svanvik, respectively. In addition, higher resolution surface ozone reanalysis for whole Europe is available from Copernicus service for regional air quality control. The regional model reanalysis ensemble mean is at fairly high spatial and temporal resolution compared to the data from ECMWF: $0.1^\circ\times 0.1^\circ$ or roughly $10.4\,\unit{km}\times 11.1\,\unit{km}$ and 1-hourly, respectively. The covered time periods differ and no data are available for before the turn of the millenium. The CAMSRA and CAMS Regional Air Quality ensemble mean are available in near realtime, but only former covers a period of sufficient length for climate analysis (2003 -- present). Their quality with respect to observations will be discussed in Section~\ref{subsubsec:clim_ozone}. Without too much foreshadowing, the results prevented us from fetching the whole CAMSRA data.

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from Svanvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged temperatures (red line) are displayed with standard error of mean as error band. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consequative months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classifcation of a regular subarctic climate (Dfc). The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the time period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

Winter and spring (November--April, except for March) are relatively dry ($\sum_m \mathrm{Precip} < 20\,\unit{mm}$). The driest month is January with $\sum_m \mathrm{Precip} = (16.7\pm 3.0)\,\unit{mm}$. The most precipitation occurres in the summer months, with a highest monthly averaged accumulated precipitation $\sum_m \mathrm{Precip} = (58.5\pm 9.2)\,\unit{mm}$ in August. The average annual accumulated precipitation given with standard error of mean is $(383\pm 86)\,\unit{mm}$. Precipitation shows a large veriablity throughout the years.

A climatology derived for global irradiance $Q_0$ is shown in Fig.~\ref{fig:global_rad_clim}a) for both, climatological average of hourly maximum $\left<Q_0^\mathrm{max}\right>$, minimum $\left<Q_0^\mathrm{min}\right>$, and mean $\left<Q_0^\mathrm{mean}\right>$ irradiance observed at Svanvik. In accordance to the geometrically derived sunshine duration shown in Fig.~\ref{fig:sunlight_fennoscandia}, $\left<Q_0^\mathrm{max}\right>$ approaches zero in December/January commonly known as polar night and reaches its maximum in June/July commonly referred to as midnight sun conditions. The highest observed $\left<Q_0^\mathrm{max}\right>$ is about $800\,\unit{W\,m^{-2}}$, while on average $\left<Q_0^\mathrm{mean}\right>$ it is about $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the climatological minimum $\left<Q_0^\mathrm{min}\right>$ lies at about $200\,\unit{W\,m^{-2}}$.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{climatediagram}
  \caption{Climate diagram following Walter and Lieth. The diagram is based on climatological data for Svanvik/Pasvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged temperatures (red line) are displayed with standard error of mean as error band. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star.}
  \label{fig:climatediagram}
\end{figure}

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies peak in spring, with peak values of about $48\,\unit{ppb}$ in late April. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. The ozone climatology laggs about 3 months behind the temperature climatology. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in case of Svanvik and Janiskoski can be explained due to the lower overall statistics.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs}
  \caption{Multi-annual mean of daily mean ozone at observation sites in Fennoscandia. The large spread in case of Svanvik and Janiskoski is due to the lower statistics. All climatologies peak in spring, with peak values in April. The annual average ozone concentration \chem{\left<[O_3]\right>} at Svanvik is $6.6\,\unit{ppb}$ lower than at the sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN).}
  \label{fig:ozone_climatology_fenoscandic_obs}
\end{figure}

The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is faily high (Fig.~\ref{fig:density_distribution}a, b), hence we can put those data series together to derive a general ozone climatology for northern Fennoscandia at similar altitudes. In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the 2 dimensional density distributions for both, the thus defined northern Fennoscandia (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}) are shown. Darker colors indicate higher prohablity to observe these values. On top of the density distributions, 10 days averages of daily mean, daily maximum, and daily minimum \chem{[O_3]} are displayed together with the $1 \sigma$ uncertainties. For the daily mean the standard error of mean is shown in addition (albeit hardly visible). From the northern Fennoscandic climatology, one can deduce several features. The maximum of the 10 days average minimum ozone concentration peaks about $20\,\unit{days}$ earlier than the average mean and average maximum concentrations. From minimum concentration to maximum concentration it takes about $8.5\,\unit{months}$ while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of the average maximum ozone concentration is at $52\,\unit{ppb}$ (late April) and the minimum of the average minimum ocurres at $15\,\unit{ppb}$ (beginning of September). The steep rise of \chem{[O_3]} in spring coincident with increasing temperatures as well as increase in global irradiance.

@@

\begin{figure*}[t]
  \includegraphics[width=12cm]{ozone_climatologies}
  \caption{Comparison of climatologies derived from observational data (EBAS, see Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}) with various ozone reanalysis products: (a) ECMWF: MACC reanalysis; (b) ECMWF: CAMS reanalysis; (c) Copernicus: Regional model reanalysis ensemble mean. As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis is in general too low and displays a erroneous seasonal cycle. The CAMS reanalysis matches better but does not reproduce the seasonality in northern Fennoscandia. The ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia, though the modelled ozone concentrations in the north are slightly to low.}
  \label{fig:ozone_climatologies}
\end{figure*}

@@

\appendixfigures  %% needs to be added in front of appendix figures

\begin{figure}[t]
  \includegraphics[width=8.3cm]{global_rad_clim}
  \caption{Observed global irradiance at Svanvik. The cyan lines indicate the $1\,\sigma$ level deducted from the climatology. (a) Climatology (hourly maxima, mean, and minima); deviation from climatology (b) 2018; (c) 2019.}
  \label{fig:global_rad_clim}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=8.3cm]{greening_season_change_Svanvik}
  \caption{Estimated shift and prolongation of greening season at Svanhovd over the past two decades based on the method and data described in \citet{GMD:Falk2019}.}
  \label{fig:greening_season_change_Svanvik}
\end{figure}
